Ensemble Stability of Biological Networks

### 2014-10-17 tex/supplement.tex
update supplement

For the case of continuous deterministic observables, the above process can be made more precise by associating a manifold $M_i$ to each observable, a directed graph, $G$, to the interdependencies, and a vector field $V$ over the space determined by taking the product of the manifolds associated to the collection of observables, satisfying these interdependencies for each observable  \cite{Deville}. For example, if we have two observables $\{x_1,x_2\}$ where the domains in which they are quantified are given by manifolds $\{M_1,M_2\}$ such that $x_1 \in M_1 \equiv \mathbb{R}^1$ and $x_2 \in M_2 \equiv \mathbb{R}^1$, a directed graph $G_X$ describing the dependencies between these observables and vector field $V$ with components $\{F_1,F_2\}$ defined on $M_1 \otimes M_2 \equiv \mathbb{R}^2$ satisfying the dependencies determined by $G_X$. If the system under consideration has the graph given in \reffigexamplesystemmodules
% \begin{center}
% \input{tex/examplehamdyngraph}
% \end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$
having connectivity $\connectivity$ equal to the number of edges of the graph, in this case $\connectivity = 3$. For a general system the directed graph $G_X$ that describes the manner in which each of the variables depends upon one another is given by the adjacency matrix $\adj(G_X)$ where
 \begin{displaymath}
   \adj(G_X)_{ij} = \left\{
     \begin{array}{ll}
       1, & F_i \hbox{ depends on } x_j\\
       0, & F_i \hbox{ does not depend on } x_j
     \end{array}.
   \right.
\end{displaymath} For the system $X = \{G_X, M_X, F_X\}$, where $M_X \equiv \{M_1,M_2\}$ and $F_X \equiv \{F_1,F_2\}$ such that $F_1$ is constrained to be a function of $x_2$ and $F_2$ is a function of both $x_1$ and $x_2$ yielding the flow equations
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2)\\
\frac{dx_2}{dt} & = F_2(x_1,x_2)
\end{align*}

@@

In order to accommodate the possibility of uncertainty in our
modeling, we will generalize our notion of a system on a network to
that of a system with random parameters.  Again, this will involve
three steps. First, we provide a measure space $S$
which represents the values over which our parameters can vary. Next,
instead of a single vector field, we consider a family of vector
fields parameterized by this space.  As before, for
each point $p \in S$, the vector field $V(p)$ must be consistent with
the system graph. Finally, we select a probability measure, $\mu$, which represents our understanding of which values of the parameters
are likely. In accord with Bayesian statistics, we might revise this
distribution as data comes in or use it to estimate parameters and error
bars from experimental results.

@@

Suppose that our Jacobian $A$ is an $n \times n$ matrix with real coefficients, connectivity $\connectivity$, and denote the proposition ``$A$ is stable'' as $\mathrm{stab}(A)$.  Then we may compute the robustness of this proposition using \ref{eq:robustness} once we have specified the probability densities $\mu$ and $\mu'$.  In general, these will depend upon the parameters of the non-linear system.  For the purpose of the current investigation, we shall make the ansatz that these are such that the distribution of entries of $A$ is (at least approximately) the uniform distribution $\mathcal{U}(-1,1)$ on the $\connectivity$-dimensional hypercube, $H^\connectivity$, of edge length $r=2$, centered about the origin. To model the perturbations, we resample the elements $k_1, k_2, \ldots k_m$ of this matrix from the same uniform distribution yielding
$$
\mu_{k_1,\ldots,k_m}(x,y) = \prod_{y \notin \{x_1, \ldots, x_m\} } \delta(x_j-y_j) \prod_{y \notin \{x_1,\cdots,x_m\}} \mathbf{1}_{[-1,1]} (y_j) \mathbf{1}_{[-1,1]} (x_j)
$$
Under these assumptions, the expression for robustness becomes a special case of \ref{eq:robustness}:
\begin{equation}\label{eq:condprobgen}
 R (\mathrm{stab}, \mu'_{k_1, \ldots, k_m}) =
  \frac{\int_{H^\connectivity} dx_1 \cdots dx_d \int_{H^m} {dx'}_{k_1} \cdots {dx'}_{k_m}\,
    \mathbf{1}_{stab} \mathbf{1}_{stab'}}
  {\int_{H^\connectivity} dx_1 \cdots dx_d  \, \mathbf{1}_{stab}(x_1 \cdots x_d)},
\end{equation}
where we made the abbreviations
\begin{align*}
\mathbf{1}_{stab} &= \mathbf{1}_{stab}(x_1 \cdots x_{k_1} \cdots x_{k_m} \cdots x_\connectivity) \\
\mathbf{1}_{stab'} &= \mathbf{1}_{stab'}(x_1 \cdots {x'}_{k_1} \cdots {x'}_{k_m}  \cdots x_d)
\end{align*}
i.e. $\mathbf{1}_{stab'}$ corresponds to replacing $x_{k_1}, \ldots x_{k_m}$ with their primed counterparts.

@@

A strongly connected component (SCC) of a graph is a maximal subset of vertices where each vertex within the subset can be reached from any other \cite{Cormen2009}. The strongly connected components of some examples of three variable systems are outlined in \reffigscc along with their adjacency matrices and hierarchy diagrams (to be defined below).
The decomposition of a digraph into strongly connected components corresponds to a block triangular decomposition of its adjacency matrix.  Say that the graph $G$ has strongly connected components $C_1, C_2, \ldots C_n$, which have been labelled in such a way that there are no links from vertices in component $C_i$ to component $C_j$ when $i < j$.  Label the vertices in such a way that $v_1, \ldots, v_{n_1}$ belong to $C_1$, $v_{n_1 + 1}, \ldots, v_{n_2}$ belong to $C_2$, etc.  Then, if we choose basis vectors corresponding to this labelling of the vertices, we will have $a_{ij} = 0$ whenever $i$ and $j$ correspond to different components and $i > j$.  This condition is equivalent to stating that the matrix is block triangular with blocks of size $n_1, n_2, \ldots$.

Corresponding to this decomposition we can construct a directed acyclic graph $\hier (G)$.  Each node of $\hier (G)$ corresponds to a strongly connected component of $G$. There is an edge from the node corresponding to component $C$ to the node corresponding to component $C'$ if and only if there exists a link from some vertex in $C$ to some vertex in $C'$ in $G$.  Because of the maximality of strongly connected components, $\hier (G)$ is acyclic.

@@

This map $\hier$ is many-to-one and so there is a large class of
operations which leaves $\hier(G)$ invariant for a given graph $G$.
For instance, we may interchange the positions of the strongly
connected components relative to each other \reffighiertransformations a.  Leaving the components fixed, we may move links between nodes in a component \reffighiertransformations b or between components, or even add or delete links \reffighiertransformations c.  As we shall see later, some of these operations leave invariant key dynamical quantities, so we may regard them as a symmetry groupoid of our system.

The relationship between $G$ and $\hier(G)$ for all $G$ with a given number of vertices suggests a heuristic method of quantifying the degree of hierarchy of a given graph and thus of the system structure it represents. The most hierarchical system is considered to be the graph corresponding to the total ordering, which for three nodes is given in \reffigscc (top panel). This graph maximizes the number of links between strongly connected components, which also implies maximizing the number of strongly connected components. The graph edit distance on a fixed number of vertices from one graph to another is defined as the minimum number of modifications of the first graph in order to transform it into the second \cite{Axenovich2011}. This distance between any given graph and the total ordering thus quantitatively represents how far a graph is from being perfectly hierarchical. In this work we take this distance to be the definition of hierarchy.

\subsection{Stability and robustness analysis of particular system ensembles}
For systems having two components, we can analytically compute the probability of stability and robustness from \ref{eq:condprobgen}. For those having three components, we can estimate these same quantities using Monte Carlo simulations. Systems of larger size can be analyzed using the symmetry properties of robustness extracted from this analysis. For two-variable systems having $2 \times 2$ Jacobian matrices, the aforementioned stability criteria result in the conditions $T < 0$ and $D >
0$ where $T$ and $D$ denote the trace and the determinant. Suppose we have a stable matrix
$$
\begin{bmatrix}
a & b \\
d & c
\end{bmatrix}
$$
where $a + c < 0$ and $ac > bd$.  For the case in which $x_1=a,\,x_2=b,\,x_3=c,\,x_4=d$ we need to compute what corresponds to $R(\mathrm{stab},\mu'_k)$ where $k=1 \ldots 4$. By symmetry, there are two cases to consider; resampling $a$ is equivalent to resampling $c$ and resampling $b$ is equivalent to resampling $d$ so we only need to explicitly compute $ R(\mathrm{stab},\mu'_1)$ and $R(\mathrm{stab},\mu'_2)$. Suppose that we resample $b$ to compute $R(\mathrm{stab},\mu'_2)$.
% \begin{strip}
% \begin{align}\label{eq:condprob}
% P\left(\begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \bigg| \begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)
% & = \frac{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable and } \begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}.
% \end{align}
The denominator of \ref{eq:condprobgen} in this case is given by
\begin{align*}
P\left(\textrm{stab } \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{\genfrac{}{}{0pt}{}{\genfrac{}{}{0pt}{}{ac>bd}{a+c<0}}{H^4}} da\,db\,dc\,dd\,1}{\int_{H^4} da\,db\,dc\,dd\,1}.
\end{align*}
Since the trace does not involve $b$, the $T<0$ condition will be satisfied automatically and we only need to examine the determinant. Thus, we have the inequalities $ac > b'd$ and $-1 < b' < 1$ in addition to the previous constraints leading to an expression for the numerator of \ref{eq:condprobgen}
\begin{align*}
P\left(\textrm{stab} \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \textrm{ and stab} \left( \begin{bmatrix}
a & b' \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{{{ac>b'd \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,db'\,1}{\int_{H^5} da\,db\,dc\,dd\,db'\,1}.
\end{align*}
The analogous equation for resampling $a$ is
\begin{align*}
P\left(\textrm{stab} \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \textrm{ and stab} \left( \begin{bmatrix}
a' & b \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{{{{a'c>bd \atop a' + c < 0} \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,da'\,1}{\int_{H^5} da\,db\,dc\,dd\,da'\,1}.
\end{align*}

Using this approach the probability of stability and of robustness for all two variable systems is given in \ref{tab:structstabmat}.

@@

This observation suggested that graph edit distance from \reffigscc (top panel), hierarchy, might provide a better characterization of dynamical robustness. \reffigrobusthierarchy shows dynamical robustness as a function of hierarchy, where the most hierarchical system has a hierarchy of 0. There is a monotonic correlation between the upper bound of robustness and hierarchy. \reffigconnectdist3D3x3 shows dynamical robustness as a function of both hierarchy and connectivity. The monotonic correlation between hierarchy and robustness is refined by an underlying correlation between robustness and connectivity analogous to that of \reffigconnectcycle3D3x3.

@@

Since the determinant of a triangular matrix equals the product of the determinants of its diagonal blocks, it follows that the characteristic polynomials factors as the product of the charactericstic polynomials of its diagonal blocks.  Hence, a block triangular matrix is stable if and only if its diagonal blocks are stable.  Note that this condition does not depend upon the entries off the diagonal (which correspond to links between strongly connected components) and does not depend upon what order the components appear.

Using these observations, we may express the robustness of a graph in
terms of the robustnesses of its strongly connected components.  If
the components $C_i$ of our graph $G$ has $v_i$ vertices, connectivity
$\connectivity_i$, and there are $\ell_{ij}$ links between components
$i$ and $j$, then there are a total of
$\sum_{i,j \in \hier(G)} \ell_{ij} + \sum_{i=1}^n \connectivity_i$
links in $G$.  If we resample a randomly chosen link in a component
$C_i$, then the probability that the system will remain stable after
resampling is $\langle R(\hbox{stab} (C_i), \mu') \rangle_1$.  If we resample a link
between components $C_i$ and $C_j$, the system will remain stable with
probability 1.  Hence, the probability that the system will remain
stable upon resampling a random link is the weighted average of these
probabilities:
\begin{align}\label{eq:sccrobustness}
\langle R(\hbox{stab} (G), \mu') \rangle_1 =
\frac{\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i \langle R(\textrm{stab}(C_i),\mu') \rangle_1}
     {\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i}.
\end{align}
For instance, if our graph is the one in \reffigscc (middle panels), then we have two connected components, one with two nodes, and one with one node.  From \ref{tab:structstabmat}, we know that the graph with two nodes has probability $0.25$ of being stable and robustness $0.62$.  The graph with one node corresponds to a $1 \times 1$ matrix, so we have probability $0.5$ of stability and robustness $0.5$.  Thus, the probability of our 3-node graph being stable is $0.5 \times 0.25 = 0.125$ and its robustness is
\[
\langle R(\textrm{stab}(G),\mu') \rangle_1 = \frac{2 + 0.5 + 4 \times 0.62}{2 + 1 + 4} = 0.714,
\]
which agrees with the value computed in \ref{tab:structstabmat} up to
sampling error.

As can be seen from \ref{eq:sccrobustness}, all that is required to fix robustenss is a set of SCC sizes and the number of links between them. This demonstrates that robustness is symmetric under the transformations shown in \reffighiertransformations a and b. Since transformations of the type \reffighiertransformations c change the number of links between connected components, this symmetry of $\hier$ is broken by \ref{eq:sccrobustness}. Nevertheless, a considerable degree of symmetry remains. For example, \ref{fig:robustnesssymmetries} shows the symmetry groupoid of robustness for the case in which we fix connected component sizes $\{2,1,1\}$ with a total of $3$ links between them.

Next, we derive an upper bound on the robustness of any graph
constructed from a given set of connected components.  Since
$\langle R(\textrm{stab}(C_i), \mu') \rangle_1 \le 1$ for all $i$, it follows that increasing $\ell_{ij}$ will
increase $\langle R(\textrm{stab}(G),\mu') \rangle_1$.  Given two connected components $C_i$ and $C_j$ with
$v_i$ and $v_j$ nodes respectively, we have a maximum of $n_i n_j$
links going from $C_i$ to $C_j$.  Hence, $\ell_{ij} \le v_i v_j$, so,
we conclude that
\begin{align}
\langle R(\textrm{stab}(G),\mu') \rangle_1 \le \frac{\sum\limits_{(i,j) \in \hier(G)} v_i v_j +
               \sum\limits_{i=1}^n \connectivity_i \langle R(C_i,\mu') \rangle_1}
              {\sum\limits_{(i,j) \in \hier(G)} v_i v_j +
               \sum\limits_{i=1}^n \connectivity_i},
\end{align}

@@

This argument also works when we resample more than one entry,
although the notation becomes more complicated.  Suppose that we
resample $m$ nodes.  Define
\begin{equation*}
M = \left\{(m_0, m_1, \ldots, m_n) \,\bigg|\,
m = \sum_{i=0}^n m_i \quad\&\quad
m_0 \le \sum_{i=1}^{n-1} \sum_{j=i+1}^n \ell_{ij} \quad\&\quad
\forall i \in \{1, \ldots, n\} \, m_i < \connectivity_i \right\}.
\end{equation*}
Then, given $(m_0, m_1, \ldots, m_n) \in M$, there are ${m \choose
m_0, m_1, \ldots. m_n}$ ways of choosing $m_i$ links from $C_i$ and
$m_0$ links between strongly connected components.  Hence, our
weighted average becomes
\begin{equation}
\langle R(\hbox{stab} (G), \mu') \rangle_m =
\frac{\sum\limits_{(m_0, m_1, \ldots m_n) \in M}
      {m \choose m_0, m_1, \ldots. m_n}
      \left(m_0 + \sum\limits_{i=1}^n m_i \langle R(\hbox{stab} (C_i), \mu') \rangle_{m_i} \right)}
     {\sum\limits_{(m_0, m_1, \ldots m_n) \in M}
      {m \choose m_0, m_1, \ldots. m_n} m}
\end{equation}

>>>>

For the case of continuous deterministic observables, the above process can be made more precise by associating a manifold $M_i$ to each observable, a directed graph, $G$, to the interdependencies, and a vector field $V$ over the space determined by taking the product of the manifolds associated to the collection of observables, satisfying these interdependencies for each observable  \cite{Deville}. For example, if we have two observables $\{x_1,x_2\}$ where the domains in which they are quantified are given by manifolds $\{M_1,M_2\}$ such that $x_1 \in M_1 \equiv \mathbb{R}^1$ and $x_2 \in M_2 \equiv \mathbb{R}^1$, a directed graph $G_X$ describing the dependencies between these observables and vector field $V$ with components $\{F_1,F_2\}$ defined on $M_1 \otimes M_2 \equiv \mathbb{R}^2$ satisfying the dependencies determined by $G_X$. If the system under consideration has the graph given in \reffigexamplesystemmodules$\,$
% \begin{center}
% \input{tex/examplehamdyngraph}
% \end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$
having connectivity $\connectivity$ equal to the number of edges of the graph, in this case $\connectivity = 3$. For a general system the directed graph $G_X$ that describes the manner in which each of the variables depends upon one another is given by the adjacency matrix $\adj(G_X)$ where
 \begin{displaymath}
   \adj(G_X)_{ij} = \left\{
     \begin{array}{ll}
       1, & F_i \hbox{ depends on } x_j\\
       0, & F_i \hbox{ does not depend on } x_j
     \end{array}.
   \right.
\end{displaymath} For the system $X = \{G_X, M_X, F_X\}$, where $M_X \equiv \{M_1,M_2\}$ and $F_X \equiv \{F_1,F_2\}$ such that $F_1$ is a function of $x_2$ alone and $F_2$ is a function of both $x_1$ and $x_2$ yielding the flow equations
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2),\\
\frac{dx_2}{dt} & = F_2(x_1,x_2).
\end{align*}

@@

In order to accommodate the possibility of uncertainty in our
modeling, we will generalize our notion of a system on a network to
that of a system with random parameters.  Again, this will involve
three steps. First, we provide a measure space $S$
which represents the values over which our parameters can vary. Next,
instead of a single vector field, we consider a family of vector
fields parameterized by this space.  As before, for
each point $p \in S$, the vector field $V(p)$ must be consistent with
the system graph. Finally, we select a probability measure $\mu$ which represents our understanding of which values of the parameters
are likely. In accord with Bayesian statistics, we might revise this
distribution as data comes in or use it to estimate parameters and error
bars from experimental results.

@@

Suppose that our Jacobian $A$ is an $n \times n$ matrix with real coefficients, connectivity $\connectivity$, and denote the proposition ``$A$ is stable'' as $\mathrm{stab}(A)$.  Then we may compute the robustness of this proposition using \ref{eq:robustness} once we have specified the probability densities $\mu$ and $\mu'$.  In general, these will depend upon the parameters of the non-linear system.  For the purpose of the current investigation, we shall make the ansatz that these are such that the distribution of entries of $A$ is (at least approximately) the uniform distribution $\mathcal{U}(-1,1)$ on the $\connectivity$-dimensional hypercube, $H^\connectivity$, of edge length $r=2$, centered about the origin. To model the perturbations, we resample the elements $k_1, k_2, \ldots k_m$ of this matrix from the same uniform distribution yielding
$$
\mu_{k_1,\ldots,k_m}(x,y) = \prod_{y \notin \{x_1, \ldots, x_m\} } \delta(x_j-y_j) \prod_{y \notin \{x_1,\cdots,x_m\}} \mathbf{1}_{[-1,1]} (y_j) \mathbf{1}_{[-1,1]} (x_j)
$$
Under these assumptions, the expression for robustness becomes a special case of \ref{eq:robustness}:
\begin{equation}\label{eq:condprobgen}
 R (\mathrm{stab}, \mu'_{k_1, \ldots, k_m}) =
  \frac{\int_{H^\connectivity} dx_1 \cdots dx_d \int_{H^m} {dx'}_{k_1} \cdots {dx'}_{k_m}\,
    \mathbf{1}_{stab} \mathbf{1}_{stab'}}
  {\int_{H^\connectivity} dx_1 \cdots dx_d  \, \mathbf{1}_{stab}(x_1 \cdots x_d)},
\end{equation}
where we made the abbreviations
\begin{align*}
\mathbf{1}_{stab} &= \mathbf{1}_{stab}(x_1 \cdots x_{k_1} \cdots x_{k_m} \cdots x_\connectivity), \\
\mathbf{1}_{stab'} &= \mathbf{1}_{stab'}(x_1 \cdots {x'}_{k_1} \cdots {x'}_{k_m}  \cdots x_d),
\end{align*}
i.e. $\mathbf{1}_{stab'}$ corresponds to replacing $x_{k_1}, \ldots x_{k_m}$ with their primed counterparts.

@@

A strongly connected component (SCC) of a graph is a maximal subset of vertices where each vertex within the subset can be reached from any other \cite{Cormen2009}. The strongly connected components of some examples of three variable systems are outlined in \reffigscc$\,$ along with their adjacency matrices and hierarchy diagrams (to be defined below).
The decomposition of a digraph into strongly connected components corresponds to a block triangular decomposition of its adjacency matrix.  Say that the graph $G$ has strongly connected components $C_1, C_2, \ldots C_n$, which have been labelled in such a way that there are no links from vertices in component $C_i$ to component $C_j$ when $i < j$.  Label the vertices in such a way that $v_1, \ldots, v_{n_1}$ belong to $C_1$, $v_{n_1 + 1}, \ldots, v_{n_2}$ belong to $C_2$, etc.  Then, if we choose basis vectors corresponding to this labelling of the vertices, we will have $a_{ij} = 0$ whenever $i$ and $j$ correspond to different components and $i > j$.  This condition is equivalent to stating that the matrix is block triangular with blocks of size $n_1, n_2, \ldots$.

Corresponding to this decomposition we can construct a directed acyclic graph $\hier (G)$ or the condensed graph \cite{Corominas-Murtra2013}.  Each node of $\hier (G)$ corresponds to a strongly connected component of $G$. There is an edge from the node corresponding to component $C$ to the node corresponding to component $C'$ if and only if there exists a link from some vertex in $C$ to some vertex in $C'$ in $G$.  Because of the maximality of strongly connected components, $\hier (G)$ is acyclic.

@@

This map $\hier$ is many-to-one and so there is a large class of
operations which leaves $\hier(G)$ invariant for a given graph $G$.
For instance, we may interchange the positions of the strongly
connected components relative to each other \reffighiertransformations$a$.  Leaving the components fixed, we may move links between nodes in a component \reffighiertransformations$b$ or between components, or even add or delete links \reffighiertransformations$c$.  As we shall see later, some of these operations leave invariant key dynamical quantities, so we may regard them as a symmetry groupoid of our system.

The relationship between $G$ and $\hier(G)$ for all $G$ with a given number of vertices suggests a heuristic method of quantifying the degree of hierarchy of a given graph and thus of the system structure it represents. The most hierarchical system is considered to be the graph corresponding to the total ordering, which for three nodes is given in \reffigscc (top panel). This graph maximizes the number of links between strongly connected components, which also implies maximizing the number of strongly connected components. The graph edit distance (ED) on a fixed number of vertices from one graph to another is defined as the minimum number of modifications of the first graph in order to transform it into the second \cite{Axenovich2011}. This distance between any given graph and the total ordering thus quantitatively represents how far a graph is from being perfectly hierarchical. In this work we take $max(ED) - ED$ to be the definition of hierarchy, where $max(ED)$ is the maximum edit distance for all graphs with a given number of nodes.

\subsection{Stability and robustness analysis of particular system ensembles}
For systems having two variables, we can analytically compute the probability of stability and robustness from \ref{eq:condprobgen}. For those having three variables, we can estimate these same quantities using Monte Carlo simulations. Systems of larger size can be analyzed using the symmetry properties of robustness extracted from this analysis. For two-variable systems having $2 \times 2$ Jacobian matrices, the aforementioned stability criteria result in the conditions $T < 0$ and $D >
0$ where $T$ and $D$ denote the trace and the determinant. Suppose we have a stable matrix
$$
\begin{bmatrix}
a & b \\
d & c
\end{bmatrix}
$$
where $a + c < 0$ and $ac > bd$.  For the case in which $x_1=a,\,x_2=b,\,x_3=c,\,x_4=d$ we need to compute what corresponds to $R(\mathrm{stab},\mu'_k)$ where $k=1 \ldots 4$. By symmetry, there are two cases to consider; resampling $a$ is equivalent to resampling $c$ and resampling $b$ is equivalent to resampling $d$ so we only need to explicitly compute $ R(\mathrm{stab},\mu'_1)$ and $R(\mathrm{stab},\mu'_2)$. Suppose that we resample $b$ to compute $R(\mathrm{stab},\mu'_2)$.
% \begin{strip}
% \begin{align}\label{eq:condprob}
% P\left(\begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \bigg| \begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)
% & = \frac{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable and } \begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}.
% \end{align}
The denominator of \ref{eq:condprobgen} in this case is given by
\begin{align*}
P\left(\textrm{stab } \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{\genfrac{}{}{0pt}{}{\genfrac{}{}{0pt}{}{ac>bd}{a+c<0}}{H^4}} da\,db\,dc\,dd\,1}{\int_{H^4} da\,db\,dc\,dd\,1}.
\end{align*}
Since the trace does not involve $b$, the $T<0$ condition will be satisfied automatically and we only need to examine the determinant. Thus, we have the inequalities $ac > b'd$ and $-1 < b' < 1$ in addition to the previous constraints leading to an expression for the numerator of \ref{eq:condprobgen}
\begin{align*}
P\left(\textrm{stab} \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \textrm{ and stab} \left( \begin{bmatrix}
a & b' \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{{{ac>b'd \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,db'\,1}{\int_{H^5} da\,db\,dc\,dd\,db'\,1}.
\end{align*}
The analogous equation for resampling $a$ is
\begin{align*}
P\left(\textrm{stab} \left( \begin{bmatrix}
a & b \\
d & c
\end{bmatrix} \right) \textrm{ and stab} \left( \begin{bmatrix}
a' & b \\
d & c
\end{bmatrix} \right) \right) = \frac{\int_{{{{a'c>bd \atop a' + c < 0} \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,da'\,1}{\int_{H^5} da\,db\,dc\,dd\,da'\,1}.
\end{align*}
Using this approach the probability of stability and of robustness for all two variable systems is given in \ref{tab:structstabmat}.

@@

This observation suggested that graph edit distance from \reffigscc (top panel), hierarchy, might provide a better characterization of dynamical robustness. \reffigrobusthierarchy$\,$ shows dynamical robustness as a function of hierarchy. There is a monotonic correlation between the upper bound of robustness and hierarchy. \reffigconnectdist3D3x3$\,$ shows dynamical robustness as a function of both hierarchy and connectivity. The monotonic correlation between hierarchy and robustness is refined by an underlying correlation between robustness and connectivity analogous to that of \reffigconnectcycle3D3x3.

@@

Since the determinant of a triangular matrix equals the product of the determinants of its diagonal blocks, it follows that the characteristic polynomial factors as the product of the charactericstic polynomials of its diagonal blocks.  Hence, a block triangular matrix is stable if and only if its diagonal blocks are stable.  Note that this condition does not depend upon the entries off the diagonal (which correspond to links between strongly connected components) and does not depend upon what order the components appear.

Using these observations, we may express the robustness of a graph in
terms of the robustnesses of its strongly connected components.  If
each component $C_i$ of our graph $G$ has $v_i$ vertices, connectivity
$\connectivity_i$, and there are $\ell_{ij}$ links between components
$i$ and $j$, then there are a total of
$\sum_{i,j \in \hier(G)} \ell_{ij} + \sum_{i=1}^n \connectivity_i$
links in $G$.  If we resample a randomly chosen link in a component
$C_i$, then the probability that the system will remain stable after
resampling is $\langle R(\hbox{stab} (C_i), \mu') \rangle_1$.  If we resample a link
between components $C_i$ and $C_j$, the system will remain stable with
probability 1.  Hence, the probability that the system will remain
stable upon resampling a random link is the weighted average of these
probabilities:
\begin{align}\label{eq:sccrobustness}
\langle R(\hbox{stab} (G), \mu') \rangle_1 =
\frac{\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i \langle R(\textrm{stab}(C_i),\mu') \rangle_1}
     {\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i}.
\end{align}
For instance, if our graph is the one in \reffigscc (middle panels), then we have two connected components, one with two nodes, and one with one node.  From \ref{tab:structstabmat}, we know that the graph with two nodes has probability $0.25$ of being stable and robustness $0.62$.  The graph with one node corresponds to a $1 \times 1$ matrix, so we have probability $0.5$ of stability and robustness $0.5$.  Thus, the probability of our 3-node graph being stable is $0.5 \times 0.25 = 0.125$ and its robustness is
\[
\langle R(\textrm{stab}(G),\mu') \rangle_1 = \frac{2 + 0.5 + 4 \times 0.62}{2 + 1 + 4} = 0.714,
\]
which agrees with the value computed in \ref{tab:structstabmat} up to
sampling error.

As can be seen from \ref{eq:sccrobustness}, all that is required to fix robustenss is a set of SCC sizes and the number of links between them. This demonstrates that robustness is symmetric under the transformations shown in \reffighiertransformations $a$ and $b$. Since transformations of the type \reffighiertransformations $c$ change the number of links between connected components, this symmetry of $\hier$ is broken by \ref{eq:sccrobustness}. Nevertheless, a considerable degree of symmetry remains. For example, \ref{fig:robustnesssymmetries} shows the symmetry groupoid of robustness for the case in which we fix connected component sizes $\{2,1,1\}$ with a total of $3$ links between them.

Next, we derive an upper bound on the robustness of any graph
constructed from a given set of connected components.  Since
$\langle R(\textrm{stab}(C_i), \mu') \rangle_1 \le 1$ for all $i$, it follows that increasing $\ell_{ij}$ will
increase $\langle R(\textrm{stab}(G),\mu') \rangle_1$.  Given two connected components $C_i$ and $C_j$ with
$v_i$ and $v_j$ nodes respectively, we have a maximum of $v_i v_j$
links going from $C_i$ to $C_j$.  Hence, $\ell_{ij} \le v_i v_j$, so,
we conclude that
\begin{align}
\langle R(\textrm{stab}(G),\mu') \rangle_1 \le \frac{\sum\limits_{(i,j) \in \hier(G)} v_i v_j +
               \sum\limits_{i=1}^n \connectivity_i \langle R(C_i,\mu') \rangle_1}
              {\sum\limits_{(i,j) \in \hier(G)} v_i v_j +
               \sum\limits_{i=1}^n \connectivity_i},
\end{align}

@@

This argument also works when we resample more than one entry,
although the notation becomes more complicated.  Suppose that we
resample $m$ nodes.  Define
\begin{equation*}
M = \left\{(m_0, m_1, \ldots, m_n) \,\bigg|\,
m = \sum_{i=0}^n m_i \quad\&\quad
m_0 \le \sum_{i=1}^{n-1} \sum_{j=i+1}^n \ell_{ij} \quad\&\quad
(\forall i \in \{1, \ldots, n\}) \; m_i < \connectivity_i \right\}.
\end{equation*}
Then, given $(m_0, m_1, \ldots, m_n) \in M$, there are ${m \choose
m_0, m_1, \ldots. m_n}$ ways of choosing $m_i$ links from $C_i$ and
$m_0$ links between strongly connected components.  Hence, our
weighted average becomes
\begin{equation}
\langle R(\hbox{stab} (G), \mu') \rangle_m =
\frac{\sum\limits_{(m_0, m_1, \ldots m_n) \in M}
      {m \choose m_0, m_1, \ldots. m_n}
      \left(m_0 + \sum\limits_{i=1}^n m_i \langle R(\hbox{stab} (C_i), \mu') \rangle_{m_i} \right)}
     {\sum\limits_{(m_0, m_1, \ldots m_n) \in M}
      {m \choose m_0, m_1, \ldots. m_n} m}
\end{equation}
